Finding the True MVP

### 2024-12-04 paper/paper.qmd
Edits to paper and data sheet
Final Edits

Basketball’s Most Valuable Player (MVP) award represents one of the most prestigious accolades in professional sports, honoring the individual deemed to have contributed the most to their team’s success over a season. While the selection process has traditionally relied on expert voting and narrative-driven narratives, the question of whether the MVP truly reflects a player's value remains open to debate. This paper seeks to address this issue by evaluating the award through a data-driven lens, introducing the MVP Index as a novel measure of player impact. By exploring historical trends, inconsistencies, and predictive modeling, this research aims to shed light on how MVP selections align—or fail to align—with statistical measures of excellence.

@@

The results of the analysis reveal several significant insights. In some seasons, the MVP Index aligns closely with the actual MVP winner, validating the selection process. However, there are also notable discrepancies, where players with superior statistical performances were overlooked in favor of candidates with compelling narratives or team success factors. These findings suggest that while data can provide clarity, the MVP award is often shaped by qualitative elements that extend beyond pure statistics.

@@

The remainder of this paper is structured as follows. @sec-data outlines the data sources, variables and processing steps that were use to help construct our very own `MVP Index`. Next, @sec-model  talks about our model and how it was developped as well as how we are using it to predict the future statistics. Then, @sec-results, we analyze MVP statistics over that nearly 40 years and how they have changed over time and highlights important discrepancies between our model and past MVPs. Lastly, @sec-discussion discusses the broader implications of these findings and acknowledges the paper's limitations.

@@

We use the statistical programming language R [@citeR], as well as the following packages to help clean and manipulate our data, [@tidyverse], [@rstanarm], [@dplyr], [@janitor], [@lubridate], [@modelsummary], [@knitr], [@StanHeaders], [@kableExtra]. Using these packages we cleaned our data, added years for the data sets that only had seasons, referring to the basketball season, as well as helped create some nice tables, unlike the traditional R software can create.

Our data was collected from [@bbref], we collected data from every player that received a vote for any All-NBA teams since 1986, this includes a vote cast for the First All-NBA Team, the Second All-NBA Team, and the Third All-NBA Team. As we believed that taking the players that received votes for the MVP awards would result in a smaller sample. We also created a new statistic call `MVP_Index` that will be defined in @sec-variables. In addition, we collected a data set for all NBA MVPs since 1986, so we could compare to the ones we have chosen with the `MVP_Index`.

Following @tellingstories, we use techniques and methods from this text we have analyzed the data using models, graphs, tables and other methods of data analysis to conduct our research to find  players that could have been MVPs and what the MVP for the following years will look like(statistically).

@@

Data collected in sports and more specifically basketball is rather simple. When an NBA game is being played there are people who's entire job is to keep track of a players statistics throughout the game, whenever a player makes a shot, or steals they ball it is being logged by that specific person. After the standard 82 games, there could be morte or less under different circumstances such ast COVID-19, injuries, or even trades all regular season statistics have been collected. As most of our in game statistics are based on the players 'per game' average, the total number of points blocks, etc., are recorded it is divided by the number of games the player played that season.

@@

Our outcome variable is the same as our estimand in @sec-intro. As `MVP Index` is dependent on our other statistics, Points, Assists, Rebounds, Steals, Blocks, and Win Shares. As we are studying MVPs in our paper this is our Variable of interest ast it represents how good a player was in comparison to the other best players in the NBA that year, from the sample of players that have received any ALL NBA Team vote.

@@

The goal of our modelling strategy is twofold. Firstly, we will be using a linear regression model to predict the MVP Index of the mvp in the following year regardless of the player. This will tell us how much better we can expect the 2025 MVP to be compared to his fellow NBA players. Secondly, We will use another data set we have created by selecting players that had the highest MVP index in their respective year whether they had won MVP or not. So, Using both these predictions we will be able to compare whether we believe the player with the highest MVP index will in fact be the next NBA MVP.

Here we briefly describe the Linear analysis model used to investigate what the MVP Index of the next NBA MVP will be. For our model we will use collected data describe in @sec-data.

@@

We expect a positive relationship between Between MVP Index and points, assists, rebounds, block, steals and win shares, as as these are all variables that define the MVP Index. In addition we expect a positive relationship as all of these statitcs indicate a good player, the larger anyone of these stats stats are the larger the MVP Index will be for the respective player. Just because a player does not aquire many block or many asists does not mean a player will have a low MVP rating, this just mean they will have to compensate with their other stats. we can also see in @sec-model-details that we have a p-value of less than $2.2 \times 10^{-16}$, which is very low and gives us confidence the the model is accurate.

@@

Below we can see @tbl-modelresults where we can see the MVPs from the last 10 years, including an additional unknown MVP with a projected MVP Index score. For this table we used both Historical Data with past MVP winners as well as data we have created on our own. The data we have created includes players that have had the highest MVP Index within their respective year, so not necessarily the same player that truly won the MVP Award that year. As we can see the historical data predicts a player to have an MVP index of 2.40 while our data suggests that the Player with the highest MVP index will be 2.45, as such it is very likely that the player with the highest MVP Index will not be the player selected for the MVP award. As with the 2024 season the player awarded with the MVP award was Nikola Jokic with an MVP Index score of 2.16, while Luka Doncic another player well deserving of the MVP Award and was in the race to win the award has a score of 2.20 which is in fact higher than Nikola Jokic's.

@@

In this section we will be taking a look at some statistics since 1986, starting with MVP index, the moving on to PPG, the RPG, and lastly we will look at APG. In these graphs we will be looking at Historical MVPs as well as players from each year with the highest MVP index. Analyzing these graphs will help us will help us understand how these stats developped through time among the best of the best in the NBA and potentially which statistics are valued more in real life than in the real MVP race.

@@

```
In @fig-mvpindex we see that both graphs seem to bounce around a lot until we reach the 2020s where both seem to converge at around an MVP Index score of 2.2. This could mean several things the MVPs are getting worse which is incredibly unlikely and we do not believe that this is the case, another possibility is that the competition is getting better and more consistent. We must remember that the MVP index relies just as much on all the other players in the data set as much as it relies on the player who's MVP Index we are looking at. If we look at the 2004-05 and the 2005-06 seasons we can see a large dip in historical MVPs, this can be attributed to Phoenix Suns' point guard Steve Nash, as he was not even in the top 10 of MVP index with two teammate Amar'e Stoudemire and Shawn Marion two of Nash's teammates both ahve a higher Index than Nash. Although these are some of the most contentious MVP selections in the past 20 years, it must be considered what NBA analysts saw in him, as his team did have success, although falling short of the championship, and he was a great leader both on and off the court, as a floor general and a mentor. The question still remains did Nash deserve those MVPs? And is the Selection process for choosing an MVP flawed?

@@

We will no be looking at @fig-pts where we are looking at MVP's points per game through the years. As this graph is similar to @fig-mvpindex in the sense that then numbers do seem to be bouncing around although here we see that they do not seem to be converging around a single point. Although in this grapgh we see something interesting as well, Both Simulated MVs and Historical MVPs there seems to be an increasing baseline as well as an increasing top line, while the points are bouncing up and down through the years there along the top data points and the bottom data points, excluding Steve Nash, the only player to win MVP while scoring under 20ppg since 1986, who we discussed in @sec-mvp-index. The interesting thing is that this occurs in both graphs, this tells us that as the years go on the points scored by MVPs, even those who tend to score less are increasing this can be confirmed in a the paper [@yan].

@@

print(trb_plot)
```
In @fig-reb we are considering at rebounds per game. This gives us some additional information on the type of MVPs and what was prioritized at which points in time. We can see that there is a significant dip between the mid 2000s and the mid 2010s when we look at Historical Data and simulated MVPs there is a lack of 'Big Men', Centers and Power Forwards, as they are often the players with the most rebounds on a given team especially until the late 2010s, when smaller players such as Russell Westbrook Began recording more more rebounds. During this time there seems to be a vacuum of elite big men, from Shaquille O'neal, Tim Duncan, and Kevin Garnett in the early 2000s to the void we see above until the late 2010s and early 2020s where Giannis Antetekumpo, Nikola Jokic, and Joel Embiid rose to power and one of the three previously mentioned players won the MVP award in the last 6 years, which is why we can see the big jump in rebounds in the last few years of the graphs.

@@

Lastly we will be looking at @fig-ast in which we graph assists per game over time since 1986. Just as in @fig-reb we have a dip in assists per game among MVPs and Simulated MVPs in which the assists per game in the 1990s and the early 2000s, where big men, while know for their rebounds and not so much for their assists were dominating the league. While we see a gradual increase in the 2020s and 2020s with the small outlier of Joel Embiid in 2023. After looking at these last few visuals we can come to the conclusion that each of these statistics seem to have a somewhat of a growing trend whether consistent or not it does seem to exists, we can come to the conclusions that MVPs in the modern NBA are more likely to be selected if they are a well rounded player, a player that score a lot of points, rebound the ball, and be an unselfish player by by passing the ball a lot especially in scoring situations.

@@

We must remember that at the end of the day the MVP is selected by a group of NBA analysts that come together once a year and cast their ballot for who they think the MVP should be. We must also remember that the award is called the 'Most Valuable Player' Award and not the 'Best Player' Award indicating that the player that should be selected is more valuable to their team than any other player. As people are deciding the player to receive these awards, it is good as they are able to take into account quantifiable and intangibles that machines would not, such as leadership, teammate morale, on-court behavior, off-court behavior and many more attributes that could affect their vote, as a machine would likely create a and MVP Index not completely unlike ours and choose the highest or lowest number depending on how it is measured.On the other hand, humans are extremely prone to bias, for example Stephen Curry was named the first unanimous  MVP in 2016 when he received all 131 votes cast that year, while in 2013 LeBron James received 120 out of 121 votes, the last vote went to Carmelo Anthony, who did not have a bad year by any means, but LeBron's Greatness that year was undeniable. Lebron's incredible statistical performance that year was one thing to behold, but he would go on to win his second NBA championship against one the most put together teams ever assembled. This shows that no matter how uncorrupt and trust worthy a set of analysts may seem they are human at the end of the day and are prone to error at the least and bias at the worst.

Sometimes choosing an mvp that would make a good story and in utn make the NBA more money also occasionally happens, for example when derrick rose won the MVP award in 2011 few people would make the statement that he was the best player in the world, as in almost every statistical category Lebron James was superior and Lebron's team, the Miami Heat, had more playoff success that year. An additional point, Derrick rose was the home town hero in Chicago, after growing in the windy city he was drafted by his hometown team, and attempted to regain the success the team had in the !990s with Michael Jordan. This makes an incredible story.

@@

In contrast, players who showcase moments of sheer brilliance often become the face of the league. These athletes might not exhibit the same game-to-game reliability, but their ability to dominate in spectacular fashion leaves a lasting impression on fans and voters alike. Peak performers can change the momentum of a season with game-winning shots, record-breaking performances, or memorable battles against top opponents. However, their inconsistency can leave teams vulnerable during less-publicized matchups or critical stretches of the season.Ultimately, the MVP conversation remains an evolving debate, shaped by both quantitative measures and subjective interpretations of what makes a player "most valuable."

@@

The MVP award voting process relies on a panel of sportswriters and broadcasters across the United States and Canada, with each voter ranking their top five players. The point system allocates 10 points for a first-place vote, 7 for second, 5 for third, 3 for fourth, and 1 for fifth. The player with the highest total points is named the MVP. While this system ensures broad coverage across media markets, it introduces potential biases rooted in the selection of voters, the weighting system, and the lack of transparency in the criteria used by individual voters.

@@

# Table By MVP Index {#sex-index}

>>>>

Basketball’s Most Valuable Player (MVP) award represents one of the most prestigious accolades in professional sports, honoring the individual deemed to have contributed the most to their team’s success over a season. While the selection process has traditionally relied on expert voting and narrative-driven considerations, the question of whether the MVP truly reflects a player's value remains open to debate. This paper seeks to address this issue by evaluating the award through a data-driven lens, introducing the MVP Index as a novel measure of player impact. By exploring historical trends, inconsistencies, and predictive modeling, this research aims to shed light on how MVP selections align—or fail to align—with statistical measures of excellence.

@@

The results of the analysis reveal several significant insights. In some seasons, the MVP Index aligns closely with the actual MVP winner, validating the selection process. However, there are also notable discrepancies where players with superior statistical performances were overlooked in favor of candidates with compelling narratives or team success factors. These findings suggest that while data can provide clarity, the MVP award is often shaped by qualitative elements that extend beyond pure statistics.

@@

The remainder of this paper is structured as follows. Section @sec-data outlines the data sources, variables, and processing steps used to construct the MVP Index. Section @sec-model details our model, explaining its development and how it is applied to predict future statistics. In Section @sec-results, we analyze MVP statistics over nearly 40 years, examining how they have evolved and highlighting important discrepancies between our model and historical MVP selections. Finally, Section @sec-discussion discusses the broader implications of these findings and acknowledges the limitations of this study.

@@

We use the statistical programming language R [@citeR] along with several packages to clean and manipulate our data, including [@tidyverse], [@rstanarm], [@dplyr], [@janitor], [@lubridate], [@modelsummary], [@knitr], [@StanHeaders], and [@kableExtra]. These tools allowed us to clean the data, add years to datasets that only included basketball seasons, and generate well-formatted tables that go beyond the basic functionality of traditional R software.

Our data was collected from [@bbref], focusing on every player who received a vote for any All-NBA Team since 1986. This includes votes for the First, Second, and Third All-NBA Teams. We chose this approach to ensure a larger and more comprehensive sample, as limiting the scope to players who received MVP votes would have resulted in a smaller dataset. Additionally, we created a new statistic called MVP_Index, which is defined in @sec-variables. This statistic was used to evaluate players and compare them to the official NBA MVPs since 1986, providing insights into alternative MVP selections.

Building on the methodologies outlined in [@tellingstories], we employed various techniques to analyze the data. These included statistical models, visualizations, tables, and other methods of data analysis to identify players who could have been MVPs and to predict what future MVPs might look like, based on their statistical profiles.

@@

Data collection in sports, and basketball in particular, is both straightforward and meticulous. Every NBA game has dedicated statisticians whose sole responsibility is to record player statistics in real-time. Whether it is a made shot, an assist, a rebound, a steal, or a block, every action is logged systematically to ensure accuracy and completeness. This process creates a detailed statistical profile for each player, capturing their performance throughout the game.

Over the course of a standard 82-game season, these statistics accumulate into a comprehensive dataset. However, the number of games played can vary due to external factors such as pandemics like COVID-19, injuries that sideline players, or trades that impact availability. Despite these variations, the collected statistics remain consistent in format and provide a foundation for deeper analysis.

@@

Our outcome variable is the same as our estimand in @sec-intro. The MVP Index is dependent on our other statistics: Points, Assists, Rebounds, Steals, Blocks, and Win Shares. As we are studying MVPs in our paper, this is our variable of interest as it represents how good a player was in comparison to the other best players in the NBA that year, from the sample of players who received any All-NBA Team vote.

@@

The goal of our modeling strategy is twofold. Firstly, we will use a linear regression model to predict the MVP Index of the MVP in the following year, regardless of the player. This will allow us to estimate how much better the 2025 MVP is likely to perform compared to his fellow NBA players. Secondly, we will utilize another dataset that we created by identifying players with the highest MVP Index in their respective years, whether they won the MVP award or not. By combining these predictions, we can assess whether the player with the highest MVP Index is likely to become the next NBA MVP.

Here, we briefly describe the linear analysis model used to predict the MVP Index of the next NBA MVP. For our model, we rely on the collected data described in @sec-data.

@@

We expect a positive relationship between MVP Index and points, assists, rebounds, blocks, steals, and win shares, as these variables collectively define the MVP Index. Additionally, we anticipate a positive relationship because all these statistics indicate strong player performance. The larger any one of these statistics is, the larger the MVP Index will be for the respective player. However, a player who does not acquire many blocks or assists will not necessarily have a low MVP rating; they will simply need to compensate with higher performance in other areas. As shown in @sec-model-details, the model has a p-value of less than $2.2 \times 10^{-16}$, which is very low and gives us confidence that the model is accurate.

@@

Below, we reference @tbl-modelresults, which presents the MVPs from the last 10 years, along with an additional unknown MVP with a projected MVP Index score. This table includes both historical data with past MVP winners and data we have created independently. Our data identifies players with the highest MVP Index within their respective years, which may not necessarily be the same players who were officially awarded the MVP that year.

As seen in the table, the historical data predicts a player to have an MVP Index of 2.40, while our data suggests that the player with the highest MVP Index will have a score of 2.45. This indicates it is likely that the player with the highest MVP Index will not be the one selected as the MVP. For example, in the 2024 season, the MVP was awarded to Nikola Jokic, who had an MVP Index score of 2.16. However, Luka Doncic, another strong contender for the award, had an MVP Index of 2.20, which was higher than Jokic's score, further highlighting this discrepancy.

@@

In this section, we examine key basketball statistics since 1986, beginning with the MVP Index, followed by points per game (PPG), rebounds per game (RPG), and assists per game (APG). The graphs will compare historical MVPs with players from each year who had the highest MVP Index. Analyzing these trends will provide insight into how these statistics have evolved over time among the NBA's elite players and may reveal which statistics are valued more in practice compared to the official MVP selection process.

@@

```
In @fig-mvpindex, we observe that both graphs fluctuate significantly until the 2020s, where they converge around an MVP Index score of 2.2. This trend could suggest several possibilities. One explanation might be that MVPs are becoming less impactful, but this is highly unlikely and not supported by the evidence. A more plausible explanation is that the competition among players is becoming stronger and more consistent over time. It is important to remember that the MVP Index depends as much on the performance of all players in the dataset as it does on the individual player being evaluated.

For example, the 2004-05 and 2005-06 seasons show a notable dip in the MVP Index for historical MVPs. This can be attributed to Phoenix Suns' point guard Steve Nash, who won back-to-back MVP awards during those seasons. Interestingly, Nash did not rank in the top 10 for MVP Index, with two of his teammates, Amar'e Stoudemire and Shawn Marion, both achieving higher Index scores. While Nash's MVP selections remain some of the most debated in the past 20 years, it is important to consider the factors NBA analysts likely valued at the time. Nash's team experienced considerable success, even though they fell short of winning a championship. Furthermore, Nash was recognized for his exceptional leadership, serving as both a floor general and a mentor.

This raises critical questions: Did Nash truly deserve those MVP awards? And, perhaps more importantly, is the selection process for determining the MVP inherently flawed?

@@

We will now examine @fig-pts, which illustrates the points per game (PPG) of MVPs over the years. Similar to @fig-mvpindex, this graph shows fluctuations in the data; however, unlike the MVP Index, the values do not appear to converge around a single point. Instead, this graph reveals an interesting trend: for both Simulated MVPs and Historical MVPs, there seems to be an increasing baseline as well as an increasing upper limit. While the data points fluctuate year to year, the overall range of PPG among MVPs is expanding, with both higher peaks and higher troughs.

A notable exception to this trend is Steve Nash, the only MVP since 1986 to win the award while scoring under 20 PPG, as discussed in @sec-mvp-index. Excluding this outlier, the graphs suggest that over time MVPs, whether simulated or historical, are scoring more points on average, even those players who traditionally score less. This observation aligns with findings from [@yan], confirming that scoring among top-tier players has been steadily increasing over the years.

@@

print(trb_plot)
```
In @fig-reb, we examine rebounds per game, providing additional insight into the type of MVPs and what was prioritized at different points in time. There is a notable dip in rebounds between the mid-2000s and the mid-2010s in both the Historical Data and Simulated MVPs. This period reflects a lack of "big men," such as centers and power forwards, who traditionally dominate rebounds on their teams. Until the late 2010s, rebounding was largely the domain of these larger players, but during this time, there seems to have been a vacuum of elite big men.

The early 2000s saw dominant rebounders like Shaquille O’Neal, Tim Duncan, and Kevin Garnett. However, their era was followed by a noticeable void until the emergence of smaller players, such as Russell Westbrook, who began recording higher rebound numbers. In the late 2010s and early 2020s, elite big men like Giannis Antetokounmpo, Nikola Jokic, and Joel Embiid rose to prominence. One of these three players has won the MVP award in each of the last six years, which accounts for the significant increase in rebounds observed in the later years of the graph.

@@

Lastly, we examine @fig-ast, which graphs assists per game over time since 1986. Similar to @fig-reb, there is a dip in assists per game among both Historical MVPs and Simulated MVPs during the 1990s and early 2000s. This period was dominated by big men, who were known more for their rebounding abilities than their passing skills. However, we see a gradual increase in assists per game during the 2010s and 2020s, with a small outlier in Joel Embiid in 2023.

After reviewing these visuals, we can conclude that each of these statistics exhibits a somewhat growing trend. While the growth may not always be consistent, it does appear to exist. This suggests that MVPs in the modern NBA are increasingly likely to be selected if they are well-rounded players, athletes who not only score points and rebound effectively but also contribute as unselfish teammates by creating scoring opportunities through their passing.

@@

We must remember that, at the end of the day, the MVP is selected by a group of NBA analysts who come together once a year to cast their ballots for the player they believe deserves the award. It is also important to note that the award is called the "Most Valuable Player" and not the "Best Player" award, emphasizing that the recipient is deemed the most valuable to their team, rather than simply the best overall player. While the human element in voting allows analysts to consider both quantifiable statistics and intangible factors, such as leadership, teammate morale, and on- and off-court behavior, this subjective process introduces room for bias. Machines, in contrast, would likely base their decisions on a purely quantitative measure, such as an MVP Index similar to ours, and choose the player with the highest score.

Human bias can sometimes lead to controversial outcomes. For example, in 2016, Stephen Curry became the first unanimous MVP, receiving all 131 votes. In contrast, in 2013, LeBron James received 120 out of 121 votes, with the remaining vote going to Carmelo Anthony. While Carmelo had a strong season, LeBron’s performance that year was exceptional, combining statistical dominance with a second NBA championship win against one of the most well-constructed teams in league history. This example underscores that even a trustworthy and well intentioned group of analysts is still human and prone to occasional error or bias.

Additionally, the selection of an MVP can sometimes reflect narrative-driven decisions that align with storytelling or commercial interests. For instance, when Derrick Rose won the MVP in 2011, few would argue he was the best player in the world. By nearly every statistical measure, LeBron James had a superior season, and his team, the Miami Heat, achieved greater playoff success. However, Rose’s story resonated with fans—he was the hometown hero of Chicago, a city looking to rekindle its basketball glory from the Michael Jordan era. This compelling narrative undoubtedly contributed to his MVP selection, illustrating how storytelling can occasionally influence outcomes in ways that deviate from purely objective criteria.

@@

In contrast, players who showcase moments of sheer brilliance often become the face of the league. These athletes might not exhibit the same game to game reliability, but their ability to dominate in spectacular fashion leaves a lasting impression on fans and voters alike. Peak performers can change the momentum of a season with game-winning shots, record-breaking performances, or memorable battles against top opponents. However, their inconsistency can leave teams vulnerable during less-publicized matchups or critical stretches of the season.Ultimately, the MVP conversation remains an evolving debate, shaped by both quantitative measures and subjective interpretations of what makes a player "most valuable."

@@

The MVP award voting process relies on a panel of sportswriters and broadcasters across the United States and Canada, with each voter ranking their top five players. The point system allocates 10 points for a first-place vote, 7 for second, 5 for third, 3 for fourth, and 1 for fifth. The player with the highest total points is named the MVP. While this system ensures broad coverage across media markets, it introduces potential biases rooted in the selection of voters, the weighting system, and the lack of transparency in the criteria used by individual voters [@schmitz].

@@

# Table By MVP Index {#sec-index}
